\chapter{Snow products}
\label{ch:snow_products}


\section{IT-Snow}
\label{sec:it_snow}

IT-SNOW is a high resolution, multi-year snow reanalysis dataset covering the whole Italian territory. 
It provides daily maps of snow water equivalent (SWE), snow depth, bulk snow density and liquid water 
content of the snowpack for a period spanning from September $1^{st}$, 2010 to August $31^{st}$, 2025. 
The spatial resolution is approximately 500 meters. IT-SNOW maps are the output of an operational chain 
developed by CIMA Research Foundation called S3M Italy.

\subsection{S3M}
\label{subsec:s3m}

S3M, also known as \textit{Snow Multidata Mapping and Modelling}, is a hydrology-oriented cryospheric 
model that simulates seasonal snow evolution through time. S3M is a raster-based model treating pixels 
as independent units, as it neglects mass and energy transport processes such as wind redistribution. The 
same set of ordinary differential equations (ODEs) is solved for each grid point by means of the Euler 
method: the hourly time-step can be adjusted by the user to meet specific application needs. All input and 
output variables are distributed. 

S3M is specifically designed for large-scale, real-time application in order to be an useful tool in 
flood forecasting frameworks. This model achieves a high level of performance by focusing on key snow evolution 
drivers via precipitation-phase partitioning and a hybrid temperature and radiation-driven melt parametrization.



\section{MODIS}
\label{sec:MODIS}

The role of Satellite Remote Sensing (SRS) in global climate monitoring has become increasingly critical, providing 
spatio-temporal observations across remote regions where conventional ground-based methods are lacking. In particular, 
those kinds of measurements are highly effective for mapping snow cover extent and duration. NASA's Moderate Resolution 
Imaging Spectroradiometer (MODIS) is a major tool in this field: this 36-band sensor operates aboard two Earth Observing 
System (EOS) satellites, namely Terra (launched in December 1999) and Aqua (launched in May 2002). This tool has been 
used in order to assess snow cover duration (SCD) trends across the Greater Alpine Region \parencite{MODIS_GAR} and to 
establish SCD climatologies across Europe. One of the latest studies on snow cover metric characteristics and trends 
focuses on the Italian territory \parencite{MODIS_ITA}, and therefore on the Italian Alps, which are the primary focus of 
this thesis.

\subsection{Dataset Generation and Gap-Filling Strategy}
\label{subsec:dataset_MODIS}
In its latest version, this Italian snow cover dataset spans from 2000 to 2025 and consists of binary maps, where a value
of zero represents to bare ground, whereas one is associated only to snow covered pixels. Available on a daily basis, these
maps feature a spatial resolution of roughly 500 meters. 

The starting point of the workline that led the authors to those maps are the raw modis datas. To correctly identify snow 
covered pixels advantage is taken on a property of snow, which is that it has very high reflectance in the visible part of 
the electromagnetic spectrum, whereas it absorbs most of the shortwave radiation. This characteristics is highlighted by 
the normalized difference snow index (NDSI), which is defined as

\begin{equation}
    \mathrm{NDSI}\,=\,\frac{\mathrm{band}4\,-\,\mathrm{band}6(7)}{\mathrm{band}4\,+\,\mathrm{band}6(7)}
    \label{eq:NDSI}
\end{equation}

where band 4 is in the visible range (in particular $0.545\,-\,0.565\,\mu \mathrm{m}$) and band 6 (or 7) is in the shortwave infrared 
(namely $1.628\,-\,1.672\,\mu \mathrm{m}$ for Terra and $2.105\,-\,2.155\,\mu \mathrm{m}$ for Aqua).



\section{Po Basin}
\label{sec:po_basin}

This snow dataset focuses on the mountainous part of the Po River District, which is an over-regional entity resulting 
from the union of the Po River Basin with some smaller ones such as Conca and Reno basins. This territory, spanning eight 
regions of Northern Italy and parts of Switzerland and France, covers almost $87000\,km^2$. Assessing the amount of water 
stored as snow in the highlands of this region is particularly important, because the Po River Basin is home to roughly 
one-third of Italy's total population and produces almost $55\%$ of Italy's hydropower. This snow product is a thirty years long 
SWE reconstruction, covering a period spanning from 1992 to 2021. For every hydrological year, SWE maps for the mountainous part 
of the Po River District are given on a daily basis during the period spanning from October $3^{rd}$ to July $1^st$, which roughly
corresponds to the period where snow coverage extends beside glacerized area. These SWE maps have a spatial resolution of 
$500\,\cross\,500\,m^2$ and are created using a hybrid modeling approach "J-snow".



